# Dimensionality Reduction, Clustering, and Unsupervised Learning

## 1 Input Data Structures and Validation

- **UL1.0** Unsupervised Learning Software should explicitly document expected
  format (types or classes) for input data, including descriptions of types or
  classes which are not accepted; for example, specification that software
  accepts only numeric inputs in `vector` or `matrix` form, or that all inputs
  must be in `data.frame` form with both column and row names.
- **UL1.1*** Unsupervised Learning Software should provide distinct
  sub-routines to assert that all input data is of the expected form, and issue
  informative error messages when incompatible data are submitted.

The following code demonstrates an example of a routine which fails to meet
this standard.

```{r hclust-fail, eval = FALSE}
d <- dist (USArrests) # example from help file for 'hclust' function
hc <- hclust (d) # okay
hc <- hclust (as.matrix (d))
```
```{r hclust-fail-fakey, eval = TRUE, echo = FALSE}
d <- dist (USArrests) # example from help file for 'hclust' function
hc <- hclust (d) # okay
hc <- tryCatch (hclust (as.matrix (d)), error = function (e) e)
print (hc)
```

The latter fails, yet issues an informative error message that clearly
indicates a failure to provide sufficient checks on the class of input data.

- **UL1.2** Unsupervised learning which uses row or column names to label
  output objects should assert than input data has non-default row or column
  names, and issue an informative message when these are not provided.

The following code provides simple examples of checks whether row and column
names appear to have generic default values.

```{r row-col-names}
x <- data.frame (matrix (1:10, ncol = 2))
x
identical (rownames (x), as.character (seq (nrow (x))))
identical (colnames (x), paste0 ("X", seq (ncol (x))))
```

Messages should be issued in both of these cases.


- **UL1.3** Unsupervised Learning Software should explicitly document any
  aspects of input data (such as row names) which are not used in model
  construction.
- **UL1.4** Unsupervised Learning Software should explicitly document whether
  input data may include missing values. 
- **UL1.5** Functions in Unsupervised Learning Software which do not admit
  input data with missing values should provide informative error messages when
  data with missing values are submitted.
- **UL1.6** Unsupervised Learning Software should document any assumptions made
  with regard to input data; for example distributional assumptions, or
  assumptions that any aspects of input data have mean values of zero.
  Implications of violations of these assumptions should be both documented and
  tested.

## 2. Pre-processing and Variable Transformation

- **UL2.0** Unsupervised Learning Software should document any transformations
  applied to input data, for example conversion of label-values to `factor`,
  and should provide ways to explicitly avoid any default transformations (with
  error or warning conditions where appropriate).
- **UL2.1** For Unsupervised Learning Software which accepts missing values in
  input data, functions should implement explicit parameters controlling the
  processing of missing values, ideally distinguishing `NA` or `NaN` values
  from `Inf` values (for example, through use of `na.omit()` and related
  functions from the `stats` package).
- **UL2.2** Unsupervised Learning Software should implement pre-processing
  routines to identify whether aspects of input data are perfectly collinear,
  notably including:
    - **RE2.2a** Perfect collinearity among predictor variables
    - **RE2.2b** Perfect collinearity between independent and dependent variables

## 3. Algorithms

## 4. Return Results

- **UL4.0** Unsupervised Learning Software should return some form of "model"
  object, generally through using or modifying existing class structures for
  model objects, or creating a new class of model objects.
- **UL4.1** Unsupervised Learning Software may enable an ability to generate
  a model object without actually fitting values. This may be useful for
  controlling batch processing of computationally intensive fitting algorithms.

### 4.1 Reporting Return Results

- **RE4.2** Model objects returned by Unsupervised Learning Software should
  implement or appropriately extend a default `print` method which provides an
  on-screen summary of model (input) parameters and methods used to generate
  results. The `print` method may also summarise statistical aspects of the
  output data or results.
- **RE4.3** Unsupervised Learning Software should also implement `summary`
  methods for model objects which should summarise the primary statistics used
  in generating the model (such as numbers of observations, parameters of
  methods applied). The `summary` method may also provide summary statistics
  from the resultant model.

## 5. Documentation

## 6. Visualization

- **UL7.0** Objects returned by Unsupervised Learning Software should have
  default `plot` methods, either through explicit implementation, extension of
  methods for existing model objects, or through ensuring default methods work
  appropriately.
- **UL7.1** Where default plot methods include labelling components of return
  objects (such as cluster labels), routines should ensure that labels are
  automatically placed to ensure readability, and/or that appropriate
  diagnostic messages are issued where readability is likely to be compromised
  (for example, through attempting to place too many labels).

## 7. Testing

