
# Regression Demonstrations

This file demonstrates the application of [rOpenSci](https://ropensci.org) 's
[standards for statistical
software](https://ropenscilabs.github.io/statistical-software-review-book/standards.html)
to one [Regression
software](https://ropenscilabs.github.io/statistical-software-review-book/standards.html#regression-and-supervised-learning)
package. These applications are not intended to represent or reflect
evaluations or assessment of the packages, and particularly not of the extent
to which they fail to meet standards. Rather, the demonstrations are intended
to highlight aspects of the software which could be productively improved by
adhering to the standards, and thereby more generally to demonstrate the
general usefulness of these standards in advancing and improving software
quality.

## [lme4](https://github.com/lme4/lme4)

```{r lib, echo = FALSE, message = FALSE}
library (lme4)
```



### [General Standards](https://ropenscilabs.github.io/statistical-software-review-book/standards.html#general-standards-for-statistical-software)

#### Documentation

- [x] **G1.0** primary reference provided

**Statistical Terminology**

- [x] **G1.1** Statistical terminology clarified and defined. 

**Function-level Documentation**

- [x] **G1.2** [`roxygen`](https://roxygen2.r-lib.org/) is (conditionally and appropriately) used to document all functions.
    - [ ] **G1.2a** Internal functions generally not appropriately documented

**Supplementary Documentation**

- [x] **G1.3** Code necessary to reproduce performance claims is provided.
- [x] **G1.4** Code necessary to compare performance with alternative implementations in other R packages is provided.

#### Input Structures

**Uni-variate (Vector) Input**

- [x] **G2.0** Expected lengths of inputs generally documented and asserted.
- [x] **G2.1** Expected types of inputs generally documented and asserted.
- [x] **G2.2** Parameters expected to be univariate do not accept multivariate input
- [ ] **G2.3** *For univariate character input:*
 - [x] **G2.3a** `match.arg()` is used to only permit expected values.
 - [ ] **G2.3b** `tolower()` is not used, and character parameters are case-dependent
- [x] **G2.4** Mechanisms provided to convert between different data types
- [x] **G2.5** `factor` type inputs appropriate documented and processed.

**Tabular Input**

- [x] **G2.6** Software accepts standard tabular inputs
- [x] **G2.7** Software appropriately converts inputs to a single defined class or type.
- [ ] **G2.8** Software does not issue diagnostic messages for type conversion
  in which information is lost or added

Main routines assume data to have a "grouping factor", yet where or not the
relevant column is a `factor` or not, or whether or not it is ordered, makes no
difference. Thus submitting data in which the "grouping factor" is, for
example, a simple integer, lead to that being interpreted as a `factor`, which
is equivalent to the addition of information describing the factor levels.
Conversely, submitting as an ordered factor makes no difference, thereby
effectively removing the information that the `factor` variable is ordered.

- [x] **G2.9** Extraction or filtering of single columns from tabular is
  appropriately performed regardless of input class.

**Missing or Undefined Values**

- [x] **G2.10**--**G2.12** Missing values appropriately handled throughout, with ability for user control
- [ ] **G2.13** No options are provided for handling undefined values, rather
  routines simply error with informative messages.

#### Output Structures

- [x] **G4.0** No local files produced, so not relvant

#### Testing 

**Test Data Sets**

- [x] **G5.0** Tests use standard data sets with known properties 
- [x] **G5.1** Data sets created within package are exported 

**Responses to Unexpected Input**

- [ ] **G5.2** The package issues numerous error and warning messages, neither
  the triggering conditions nor values of which are explicitly tested.
    - [x] **G5.2a** Messages produced by `stop()`, `warning()`, `message()`are unique.
    - [ ] **G5.2b** Tests do not trigger every one of those messages
- [x] **G5.3** Absence of missing values form return results is explicitly tested.

**Algorithm Tests**

- [x] **G5.4** Correctness tests are implemented.
- [x] **G5.5** Correctness tests are run with a fixed random seed
- [x] **G5.6** Parameter recovery tests impelemented
    - [ ] **G5.6a** Many, but not all, parameter recovery tests are expected to
      succeed within a defined tolerance 
    - [ ] **G5.6b** Parameter recovery are generally only run with one single
      random seeds when algorithm contains a random component.
- [ ] **G5.7** There appear to be no algorithm performance tests
- [x] **G5.8** Edge condition appropriately tested
- [ ] **G5.9** There appear to be no noise susceptibility tests, neither tests
  of the effect of adding trivial noise, nor tests running under different
  random seeds.

**Extended tests**

- [x] **G5.10**--**G5.12** The `lme4` package provides exemplary use of exactly
  such an environmental variable, carefully documented in a `tests/README.md`
  file.


---

### [Regression and Supervised Learning Standards](https://ropenscilabs.github.io/statistical-software-review-book/standards.html#regression-and-supervised-learning)

#### Input data structures and validation

- [x] **RE1.0** Models are specified via a formula interface
- [x] **RE1.1** Documentation is provided for how formula interfaces are
      converted to matrix representations of input data. 

The main [package
vignette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)
provides a particularly exemplary demonstration (in Section 2.3) of the
relationship between the package's formula interface and the underlying matrix
representations.

- [ ] **RE1.2** Expected input formats are documented, but errors from other formats are not informative.

Documentation of the primary function (`lmer`) states that the main `data`
parameter is, "*an optional data frame containing the variables named in
`formula`*". This function fails with equivalent `matrix` input with the
uninformative error,
```{r lmer1, eval = FALSE}
s <- sleepstudy
s$Subject <- as.integer (s$Subject)
s <- as.matrix (s)
m <- lmer(Reaction ~ Days + (Days | Subject), data = s)
```

    ## Error in list2env(data) : first argument must be a named list

The function nevertheless accepts any generic rectangular input,
including `tibble`, and `data.table` formats. Rectification to this standard
would require only (i) updating the documentation to explicate that the
function accepts any objects able to be coerced to `data.frame` representation;
and (ii) ensuring that passing non-compliant `data` objects generates
informative messages.


- [x] **RE1.3** All relevant aspects of input data are transferred to return objects
- [x] **RE1.4** Assumptions made with regard to input data are clearly documented.

#### Pre-processing and Variable Transformation

- [ ] **RE2.0** Not all transformations applied to input data are documented

`lme4` implicitly converts all columns of input data specified in a model
formula as grouping factors (through being on the right-side of a vertical bar,
`|`, in the formula) to `factor`, yet this is not explicitly stated in
documentation. Avoiding such conversion would make no sense here.

- [x] **RE2.1**--**RE2.2** Appropriate processing of missing values is implemented.

`lme4` provides exemplary handling of this case, as illustrated by the following code:
```{r lmer2}
set.seed (1)
s <- sleepstudy
s$Reaction [ceiling (runif (1, max = nrow (s)))] <- NA # random NA value in response variable
m <- lmer(Reaction ~ Days + (Days | Subject), data = s)
nobs (m)
length (predict (m))
length (predict (m, s))
```

- [x] **RE2.3** Centering of data is appropriately diagnosed and implemented.
- [ ] **RE2.4** *Regression Software should implement pre-processing routines to identify whether aspects of input data are perfectly collinear, notably including:*
 - [ ] **RE2.4a** *Perfect collinearity among predictor variables*
 - [ ] **RE2.4b** *Perfect collinearity between independent and dependent variables* 

These conditions are neither pre-identified nor appropriately processed, with the
first case returning an empty model, and the second initially issuing an
appropriate message ("dropping 1 column / coefficient"), yet failing to
subsequently fit an appropriate model. The following code demonstrates:

```{r lmer3}
s1 <- s2 <- sleepstudy
s1$Reaction <- s1$Days
lmer(Reaction ~ Days + (Days | Subject), data = s1)
s2$Days2 <- 2 * s2$Days
m <- lmer(Reaction ~ Days + Days2 + (Days | Subject) + (Days2 | Subject), data = s2)
```

#### Algorithms

- [x] **RE3.0**--**RE3.3** Control over algorithmic convergence in `lme4` is
  exemplarily handled through the `lmerControl()` function and associated
  extensive documentation.


#### Return Results

- [ ] **RE4.0**--**RE4.13** All accessor methods for model data provided by the
  `stats` package, and all model parameters, are also implemented for `lmerMod`
  objects.

**Prediction, Extrapolation, and Forecasting**

- [ ] **RE4.14** *Where possible, values should also be provided for extrapolation or forecast *errors*.*
- [ ] **RE4.15** *Sufficient documentation and/or testing should be provided to demonstrate that forecast errors, confidence intervals, or equivalent values increase with forecast horizons.* 

The `lme4` package does not explicitly implement forecasting or extrapolation
algorithms, rather it is a generic model fitting package, the results of which
can be used with methods such as `predict()` to generate forecast values beyond
the ranges of input data. There are nevertheless no explicit methods to use
a model to generate confidence intervals on forecasts in the ways offered by
the `stats` package:

```{r predict1}
x <- data.frame (x = 1:10,
                 y = runif (10))
m0 <- lm (y ~ x, data = x)
predict (m0, newdata = data.frame (x = 11:16), se.fit = TRUE)
predict (m0, newdata = data.frame (x = 11:16), interval = "confidence")
predict (m0, newdata = data.frame (x = 11:16), interval = "prediction")
```
The three calls to `precict()` illustrate different ways of using the model to
generate estimates of uncertainty involved in using that model to make
forecasts. No such equivalent methods exist for objects of class `lmerMod`:


```{r predict2}
m <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)
predict (m, data.frame (Days = 0:12, Subject = 308))
predict (m, data.frame (Days = 0:12, Subject = 308), se.fit = TRUE)
predict (m, data.frame (Days = 0:12, Subject = 308), interval = "confidence")
predict (m, data.frame (Days = 0:12, Subject = 308), interval = "prediction")
```

Similar comments apply to objects of class `glmerMod` returned by the `glmer` function:

```{r predict-glmer}
gm <- glmer(cbind(incidence, size - incidence) ~ period + (1 |herd),
            data = cbpp,
            family = binomial)
p <- predict (gm, type = "response", se.fit = TRUE)
p <- predict (gm, type = "response", interval = "confidence")
p <- predict (gm, type = "response", interval = "prediction")
p # numeric values only
```


- [x] **RE4.16** It is possible to submit new groups to `predict()` methods.* 

**Reporting Return Results**

- [x] **RE4.17**--**RE4.18** Default `print` and `summary` methods for model objects are implemented.

#### Documentation

- [ ] **RE5.0** `lme4` does not explicitly document scaling of model fitting algorithms 

#### Visualization

- [x] **RE6.0** Default plot methods provided for model objects
- [x] **RE6.3** It is not possible to use `lme4` to distinguish *extrapolated*
      from *interpolated* predicted values, and thus no visual distinction is
      possible. See preceding comments regarding prediction methods under 4.2.

#### Testing

**Input Data**

- [ ] **RE7.0** *Tests with noiseless, exact relationships between predictor (independent) data.*
 - [ ] **RE7.0a** In particular, these tests should confirm ability to reject perfectly noiseless input data.
- [ ] **RE7.1** *Tests with noiseless, exact relationships between predictor (independent) and response (dependent) data.*
 - [ ] **RE7.1a** *In particular, these tests should confirm that model fitting is at least as fast or (preferably) faster than testing with equivalent noisy data (see RE2.4b).* 

These tests do not appear to explicitly exist, although the test suite is quite
large, and they could nevertheless be present somewhere.

**Diagnostic Messages**

- [ ] **RE7.2** Not all error and warning messages are explicitly triggered in tests

**Return Results**

- [x] **RE7.3** Tests demonstrate that output objects retain aspects of input data such as row or case names.
- [x] **RE7.4** Tests demonstrate expected behaviour when return objects are
      submitted to the accessor methods of **RE4.2**--**RE4.7**.
- [ ] **RE7.5** Test do not demonstrate that forecast errors, confidence intervals, or equivalent values increase with forecast horizons, although that could readily be argued to be inappropriate here.
